Canopy leaf area of a mature evergreen *Eucalyptus* woodland does not respond to elevated atmospheric [CO~2~] but tracks water availability

### 2015-08-17 manuscript.Rmd
Included litter basket table in response to reviewer.

## Leaf litter production
In each ring, eight litter cloth baskets with an area of 0.1979 m^2^ were installed at 0.5m above the forest floor. Litter was collected monthly since October 2012, and leaves were separated from other components such as twigs, bark, seeds, insect frass, etc. Litter production was converted to units of LAI using an estimate of specific leaf area (SLA) determined from 329 live, non-senescent leaves collected from the tree canopies from May 2012 to May 2014 (`r get_sla()` cm^2^ g^-1^, SE=2.14). We did not find a significant effect of C~a~ treatment on SLA (Ellsworth, unpublished data) at any of the collection dates.

@@

where $\tau_d$ is diffuse canopy transmittance, Q~above~ is above canopy PAR (µmol m^-2^s^-1^), and Q~below~ is the average PAR across the three sub-canopy PAR sensors in each FACE ring. We then calculated daily averages of $\tau_d$, excluding days where fewer than 4 time steps (i.e. 2 hours) were available, and days where the standard deviation among the three PAR sensors in a ring was > `r .maxSD` (in units of $\tau_d$).

>>>>

## Leaf litter production
In each ring, eight litter cloth baskets with an area of 0.1979 m^2^ were installed at random locations (**can I say this?**) at 0.5m above the forest floor. Litter was collected monthly since October 2012, and leaves were separated from other components such as twigs, bark, seeds, insect frass, etc. Litter production was converted to units of LAI using an estimate of specific leaf area (SLA) determined from 329 live, non-senescent leaves collected from the tree canopies from May 2012 to May 2014 (`r get_sla()` cm^2^ g^-1^, SE=2.14). We did not find a significant effect of C~a~ treatment on SLA (Ellsworth, unpublished data) at any of the collection dates.

@@

where $\tau_d$ is diffuse canopy transmittance, Q~above~ is above canopy PAR ($\mu$mol m^-2^s^-1^), and Q~below~ is the average PAR across the three sub-canopy PAR sensors in each FACE ring. We then calculated daily averages of $\tau_d$, excluding days where fewer than 4 time steps (i.e. 2 hours) were available, and days where the standard deviation among the three PAR sensors in a ring was > `r .maxSD` (in units of $\tau_d$).

### 2015-08-17 review1.Rmd
Moved LAI-BA to SuppInfo; fixed SuppInfo so it is standalone; changed Fig 2 to LAI instead of LAI anomaly; changed fig labels to LAI
Included litter basket table in response to reviewer.

In this manuscript, Duursma et al. present findings on the response of LAI and litter production to elevated atmospheric CO2 in EucFACE, a free air CO2 enrichment experiment established in a Eucalyptus woodland. The authors had expected elevated CO2 to increase LAI and litter production. In contrast to these expectations, the authors found CO2 to have no significant effect on either LAI or litter production over the 28-month experiment. However, they did find that CO2 extended leaf lifespan by approximately two months.

@@

**We now use pre-treatment BA as a covariate in a mixed-effects model. We found no CO2 or CO2 x time effect on LAI when BA was used as a covariate. There is also no effect of CO2 when BA is not included as a covariate, but due to the tight relationship between BA and LAI, using BA as a covariate increases the ability to find a CO2 effect. We only show the results when including BA as a covariate in the manuscript.**

@@

**As noted above, we have redone the analysis and find no CO2 or CO2 x time effect. Although there was a brief period when the 95% confidence intervals were not overlapping, this brief period did not result in a significant overall CO2 x time effect. For this reason we have not removed the insect attack from the study (it would have been appropriate if we did find a CO2 x time effect).**

@@

*TODO* **We have included more background on the design of the litter collection methods. As the table below shows, the number and size (and fractional area of the plot covered) are in the upper range of six other FACE experiments. It is clear from this comparison that our methods are justified. Further, the standard errors for litter within a plot BLAH BLAH**

@@

**We have moved this analysis, including the methods, to the Supporting Information. We believe this is interesting information for others concerned with LAI estimation, but agree it is a side-objective. We only briefly mention in the Methods that the two methods gave comparable results.**

Finally, the manuscript requires extensive editing for structure and flow. The introduction does not sufficiently establish the expectation of litter responses.

@@

**We have done so in response to specific comments listed below**

@@

**We have responded to this issue above.**

@@

>>>>

---
output: word_document
---
In this manuscript, Duursma et al. present findings on the response of LAI and litter production to elevated atmospheric CO2 in EucFACE, a free air CO2 enrichment experiment established in a Eucalyptus woodland. The authors had expected elevated CO2 to increase LAI and litter production. In contrast to these expectations, the authors found CO2 to have no significant effect on either LAI or litter production over the 28-month experiment. However, they did find that CO2 extended leaf lifespan by approximately two months.

@@

**We now use pre-treatment BA as a covariate in a mixed-effects model. We found no CO2 or CO2 x time effect on LAI when BA was used as a covariate. There is also no effect of CO2 when BA is not included as a covariate, but due to the tight relationship between BA and LAI, using BA as a covariate increases the ability to find a CO2 effect. We only show the results when including BA as a covariate in the manuscript. Fig.2 now shows the LAI timeseries instead of the anomaly as before, and we have amended the text in various places to account for these changes.**

@@

**As noted above, we have redone the analysis and find no CO2 or CO2 x time effect. Although there was a brief period when the 95% confidence intervals were not overlapping (for the anomaly, though not when LAI is considered, see new Fig.2), this brief period did not result in a significant overall CO2 x time effect. For this reason we have not removed the insect attack from the study (it would have been appropriate if we did find a CO2 x time effect).**

@@

**and**

**The size, number, and relative coverage of the litterfall traps is broadly consistent with other forest FACE studies (see Table below), with the largest trap size. The size and coverage of the traps is therefore fully justified. The litter traps were placed in somewhat random locations across the ring, but never within XXX meters of the plot edge, to minimize contamination. We, as in all other FACE sites, cannot be completely certain that contamination did not occur, but it is likely that the vast majority of litter in each basket was produced in that FACE ring.**

```{r echo=FALSE}
tabl <- read.csv(text="Experiment,	Reference,	Nr Traps per plot,	Trap size (m2),	Plot size (m2),	Coverage (%)
Duke,Finzi et al. 2001,	12,	0.16,	908,	0.0021
Oak Ridge,Norby et al. 2003, 	7,	0.19,	491,	0.0027
PopFACE,Lagomarsino et al. 2013,	18,	0.13,	300, 0.0078
Rhinelander,Liu et al. 2009,	18,	0.145,	707,	0.0037
BangorFACE,Godbold et al. 2014,	7,	0.11,	57,	0.0136
EucFACE,this paper,	8,	0.1979,	491,	0.0032", header=TRUE, check.names=FALSE)

tabl <- tabl[,-ncol(tabl)]

library(knitr)
kable(tabl)
```

@@

**We have moved this analysis, including the methods, to the Supporting Information. We believe this is interesting information for others concerned with LAI estimation, but agree it is a secondary objective only. We only briefly mention in the Methods that the two methods gave comparable results.**

Finally, the manuscript requires extensive editing for structure and flow.

**We have made substantial edits to the text, in particular accounting for the reviewer's comments listed below.**

The introduction does not sufficiently establish the expectation of litter responses.

@@

**We have done so in response to specific comments listed below.**

@@

@@

**Done.**
